Deciding Bitvector Termination with Program Synthesis

### 2014-07-01 papers/termination/paper.tex
Add #iterations to results table
Reran benchmarks
added stuff to sec 5

\begin{definition}[Conditional Termination Formula {\bf [CT]}]
\label{def:conditional-termination-formula}
 \begin{align*}
  \exists R, Inv . \forall \mathcal{V}, \mathcal{V'} . & I(\mathcal{V}) \wedge G(\mathcal{V}) \rightarrow Inv(\mathcal{V}) ~ \wedge \\
                                 & G(\mathcal{V}) \wedge V(\mathcal{V}) \wedge T(\mathcal{V}, \mathcal{V'}) \rightarrow Inv(\mathcal{V'}) \wedge R(\mathcal{V}) > 0 
  \wedge R(\mathcal{V}) > R(\mathcal{V'})
 \end{align*}
\end{definition}

@@

@@

In order to formulate the non-termination problem inside the synthesis fragment, 
we make use of the notion of \emph{closed recurrence set} introduced by Chen et al. in \cite{DBLP:conf/tacas/ChenCFNO14}. 
As opposed to a closed recurrence set, this is a  purely universal property: for each state in the recurrence set $N$, each of its successors 
must be in $N$. Thus, the non-termination problem becomes equivalent to the
satisfiability of Definition~\ref{def:deterministic-nonterm-formula}.

\todo{non-determinism?}\\

According to \cite{DBLP:conf/popl/GuptaHMRX08,DBLP:conf/tacas/ChenCFNO14}, a transition relation is non-terminating iff it has a recurrence set of states.
Candidate transition relations are 
\emph{lassos} in the program's CFG. %(they  make use of the syntactic notion of lasso).
A lasso at a cutpoint location consists of two sequences
of transitions, the stem and the loop.
The stem is a finite path from an initial location to the cutpoint location. 
The loop is a finite path that starts and ends at the cutpoint
location by following a cyclic path through the CFG. 
A lasso induces an infinite execution if the infinite path obtained by traversing the stem and then unrolling
the loop infinitely many times is feasible. In this case, the lasso is said to be non-terminating.
%% We observe that in the case that $T$ is deterministic (that is, each state has
%% exactly one successor),  the non-termination problem is equivalent to the
%% satisfiability of Definition~\ref{def:deterministic-nonterm-formula}.

The method is incomplete as not all non-terminating
program executions are induced by lassos. An infinite behaviour
that is not periodic is illustrated by the following program:

Our method is purely semantic. %We extend the notion of recurrence set in order to formulate the 
We formulate the non-termination problem at the loop level.%, rather than transition relation.
In our design, a loop $L=(\mathcal{V},I,G,t)$ is non-terminating iff it has a recurrence set of states.\\

@@

@@

\begin{figure*}
\centering
\begin{tabular}{|l|r|c|c|c|c|c|r|r|c|r|}
\hline
    & LOC & Terminates? & \shortstack{Linear \\ program?} & \shortstack{Linear ranking \\ function?}  & Conditional? & Float? & Dimension & \shortstack{Ranking \\ program size} & Res & Time (s)\\
    \hline
    \hline
\input{table}
    \hline
\end{tabular}

>>>>

\begin{definition}[Conditional Termination Formula {\bf [CT]}]
\label{def:conditional-termination-formula}
 \begin{align*}
  \exists R, Inv . \forall \mathcal{V}, \mathcal{V'} . & I(\mathcal{V}) \wedge G(\mathcal{V}) \rightarrow Inv(\mathcal{V}) ~ \wedge \\
                                 & G(\mathcal{V}) \wedge Inv(\mathcal{V}) \wedge T(\mathcal{V}, \mathcal{V'}) \rightarrow Inv(\mathcal{V'}) \wedge R(\mathcal{V}) > 0 
  \wedge R(\mathcal{V}) > R(\mathcal{V'})
 \end{align*}
\end{definition}

@@

\begin{definition}[ Nested Loops Termination Formula {\bf [NT]}]
\label{def:deterministic-nonterm-formula}
 \begin{align*}
  \exists R, S . \forall \mathcal{V}, \mathcal{V}' . & G_1(\mathcal{V}) \wedge P_1(\mathcal{V},\mathcal{V}') \rightarrow S(\mathcal{V}',\mathcal{V}') ~ \wedge \\
                                & G_2(\mathcal{V}') \wedge S(\mathcal{V},\mathcal{V}') \wedge T(\mathcal{V}',\mathcal{V}'')\rightarrow S(\mathcal{V},\mathcal{V}'') ~ \wedge \\
				& G_1(\mathcal{V}) \wedge P_1(\mathcal{V},\mathcal{V}') \wedge \neg G_2(\mathcal{V}') \wedge S(\mathcal{V}',\mathcal{V}'') \wedge 
                                P_2(\mathcal{V}'',\mathcal{V}''') \wedge 
                                  R(\mathcal{V}) > 0 \wedge R(\mathcal{V}) > R(\mathcal{V}''') 
 \end{align*} 
\end{definition}

@@

In order to formulate the non-termination problem inside the synthesis fragment, 
we make use of the notion of \emph{closed recurrence set} introduced by Chen et al. in \cite{DBLP:conf/tacas/ChenCFNO14}. 
As opposed to a closed recurrence set, this is a  purely universal property: for each state in the recurrence set $N$, each of its successors 
must be in $N$. Thus, the non-termination problem becomes equivalent to the
satisfiability of Definition~\ref{def:deterministic-nonterm-formula}.
As our formulation is with respect to a given loop $L=(\mathcal{V},I,G,t)$, we require the  
recurrence set to include the loop's guard, $N(\mathcal{V}) \rightarrow G(\mathcal{V})$. In other words, the loop's guard needs to be satisfied in order for the 
infinite execution to happen.

Given that our state space is finite, a transition relation induces a infinite execution if and only if some state is 
visited infinitely many times in the course of the execution. Hence, an alternative formulation of the non-termination problem 
for bit-vectors may search for singleton recurrence sets (the recurrence set contains only the state visited infinitely many times).
However, this would require reasoning about the transitive closure of the transition relation.\\

\noindent {\bf Non-determinism.} A non-termination diagnosis procedure searching for a closed recurrence set under-approximates the transition relation, i.e. 
the closed recurrence set together with the under-approximation characterise non-termination \cite{DBLP:conf/tacas/ChenCFNO14}.
A potential difficulty with such a procedure lies in the handling of non-deterministic assignments. For illustration, consider the following loop:

\begin{lstlisting}[language=C]
while(x != 0) {
  y = nondet();
  x = x-y;
}
\end{lstlisting}

Only for certain values of $y$ the loop is non-terminating.
In order to restrict the values that may be assigned to $y$, we replace the non-deterministic choice with a fresh variable $a$ and use formula {\bf [CNT]}
to compute a closed recurrence set $N[a]$ parametrised by $a$. Now $N[a]$ becomes an assertion over $\mathcal{V} \cup \{a\}$.
%, i.e. we under-approximate the transition relation. 
For this example, our technique computes $N[a](x,y,a) = $.%, where $a$ corresponds to the non-deterministic choice.  \\
%The issue with the closed recurrence set is that it yields an incomplete  for loops 
%that have non-deterministic assignments 
%parametric recurrence set

%% According to \cite{DBLP:conf/popl/GuptaHMRX08,DBLP:conf/tacas/ChenCFNO14}, a transition relation is non-terminating iff it has a recurrence set of states.
%% Candidate transition relations are 
%% \emph{lassos} in the program's CFG.
%% A lasso at a cutpoint location consists of two sequences
%% of transitions, the stem and the loop.
%% The stem is a finite path from an initial location to the cutpoint location. 
%% The loop is a finite path that starts and ends at the cutpoint
%% location by following a cyclic path through the CFG. 
%% A lasso induces an infinite execution if the infinite path obtained by traversing the stem and then unrolling
%% the loop infinitely many times is feasible. In this case, the lasso is said to be non-terminating.
%% We observe that in the case that $T$ is deterministic (that is, each state has
%% exactly one successor),  the non-termination problem is equivalent to the
%% satisfiability of Definition~\ref{def:deterministic-nonterm-formula}.

%% The method is incomplete as not all non-terminating
%% program executions are induced by lassos. An infinite behaviour
%% that is not periodic is illustrated by the following program:

%Our method is purely semantic.

@@

{\bf Nested loops.}

@@

\begin{figure*}
\centering
\small
\begin{tabular}{|l|r|c|c|c|c|c|r|r|c|r|r|}
\hline
    & LOC & Terminates? & \shortstack{Linear \\ program?} & \shortstack{Linear ranking \\ function?}  & Conditional? & Float? & Dimension & \shortstack{Ranking \\ program size} & Res & Time (s) & Iterations\\
    \hline
    \hline
\input{table}
    \hline
\end{tabular}
